Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-06-27 paper/main.tex
reformat and add plots

\paragraph{TODO}
\begin{itemize}
    \item Change citation style to \textbf{[1]}, \textbf{[2]}, \dots
    \item Re-run numerical experiments with simplified and improved implementation (test pinvh against lstsq)
    \item Check that everywhere correct/consistent orders $\mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1})$)
    \item 
\end{itemize}
}

@@

In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been proposed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smooth spectral density

@@

@@

\paragraph{Contribution} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices. In particular, we reuse the same randomization for each value of the parameter $t$ in the estimators, which makes them scale favorably with the number of parameter evaluations. Subsequently, we propose a simple, fast, robust, and provable procedure in which these estimators are employed for approximating the smooth spectral density $\phi_{\sigma}$ of a matrix.

@@

\section{Analysis of trace estimators for parameter-dependent matrices}
\label{sec:analysis}

@@

\paragraph{Girard-Hutchinson estimator} We can estimate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
\begin{equation}
    \Hutch{\mtx{\Psi}}(\mtx{B}(t))
    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi})
    \label{equ:hutchinson-trace-estimator}
\end{equation}
where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere. However, our theoretical developments only hold in the Gaussian case.

@@

    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm. Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Following \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable. Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:hutchinson-trace-onequery-uniform}
    \end{equation}

@@

    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
    \begin{equation}
        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log(\delta^{-1}) \rceil}}
        = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
        \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
    \end{equation}
    from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^{\sfrac{1}{2}} \varepsilon \log(\delta^{-1}))^2$.
    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
    %\end{equation}
\end{proof}

@@

\begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}, p \geq 2$ 
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]
        \leq  (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
    \end{equation}
    %\begin{equation}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
    %\end{equation}
    %If we choose 
\end{lemma}

\begin{remark}
    In particular, this result can also be used to bound
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
    \end{equation}
    for any $k,p \in \mathbb{N}, p \geq 2$.
\end{remark}

@@

\begin{lemma}{Spectral norm moments of non-standard Gaussian random matrix}{spectral-norm-moment}
    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}, p \geq 2$ 
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        \leq  (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
    \end{equation}
    %\begin{equation}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
    %\end{equation}
    %If we choose 
\end{lemma}

\begin{remark}
    In particular, this result can also be used to bound
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\sqrt{3} \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
    \end{equation}
    for any $k,p \in \mathbb{N}, p \geq 2$. Indeed, for $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
\end{remark}

\begin{proof}
    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that for $p \geq 2$
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
    \end{equation}
    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$. Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
    \end{align}

@@

        &= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
        &\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _F \leq \sqrt{k} \lVert \mtx{A} \rVert _2$)} \notag \\

@@

        \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]

>>>>

\paragraph{TODO}
\begin{itemize}
    \item Check that everywhere correct/consistent orders $\mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1})$)
    \item 
\end{itemize}
}

@@

In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been proposed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density

@@

    \label{equ:smoothed-spectral-density}

@@

\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the algorithms from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials, and realize that an efficient implementation of all of their methods can be summarized in a single algorithm. Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.

@@

\section{Analysis of randomized trace estimators applied to parameter-dependent matrices}
\label{sec:analysis}

@@

\paragraph{Girard-Hutchinson estimator} We can approximate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ stochastically independent standard Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
\begin{equation}
    \Hutch{\mtx{\Psi}}(\mtx{B}(t))
    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi})
    \label{equ:hutchinson-trace-estimator}
\end{equation}
where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. Other choices for the distribution of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere. However, our theoretical developments only hold in the Gaussian case.

@@

    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm. Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Following \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable. Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:hutchinson-trace-onequery-uniform}
    \end{equation}

@@

    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
    \begin{equation}
        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log(\delta^{-1}) \rceil}}
        = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
        \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
    \end{equation}
    from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
    %\end{equation}
\end{proof}

@@

\begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]
        \leq  (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
    \end{equation}
    %\begin{equation}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
    %\end{equation}
    %If we choose 
\end{lemma}

\begin{remark}
    In particular, this result can also be used to bound
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
    \end{equation}
    for any $k,p \in \mathbb{N}$.
\end{remark}

@@

\begin{lemma}{Spectral norm moments of non-standard Gaussian random matrix}{spectral-norm-moment}
    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        \leq  (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
    \end{equation}
    %\begin{equation}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
    %\end{equation}
    %If we choose 
\end{lemma}

\begin{remark}
    In particular, this result can also be used to bound
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2 \right] = \sqrt{\mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]} \leq \sqrt{k + p} \cdot \left(\sqrt{3} \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F\right)
    \end{equation}
    for any $k,p \in \mathbb{N}$. Indeed, for $p=2$ this bound is slightly less tight than its equivalent in \cite[lemma B.1]{tropp-2023-randomized-algorithms}.
\end{remark}

\begin{proof}
    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
    \end{equation}
    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$. Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
    \end{align}

@@

        %&= \mathbb{E}^{p}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
        %&\leq \mathbb{E}^{p}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
        &\leq \mathbb{E}^{p}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\

@@

        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \right]
